Computational design of RNA-based oscillatory circuits

### 2015-09-03 MScWriteup.tex
more writing

 We can test the predictive ability of our model by cross validating - taking the parameter values found in the fitting of one dataset, and using them to give model predictions for another. We thus take the parameter values giving the best fit for the  \textit{13\_9} dataset and use them to predict the \textit{14\_7} dataset, and vice versa. Results are shown in Fig.\ref{CrossValidation}.Though the predictions are reasonable, they are substantially worse than the predictions for the data they were trained on. The first reason for this may be experiment to experiment variability of parameter values \cite{Hu2015}. The second may be that the model is overfitting.

@@

 We may investigate whether our model suffers from such problems by performing a local sensitivity analysis about one of the solutions found in our initial parameter estimation. We numerically estimate the sensitivity matrix, $S$:

@@

It can be shown that if sensitivity co-efficients are linearly dependent over the range of observation values, the associated parameters cannot be simultaneously estimated \cite{Beck}. Related to this fact, a number of measures have been proposed to asses parameter estimability from the sensitivity matrix \cite{Mclean2012}

Fig. \ref{SensitivityMatrix} shows plots of each column of $S$, evaluated at the parameter set giving the lowest error in the \textit{13\_9} dataset - these show the evolution of the sensitivity matrix over time. Fig. \ref{SensitivityMatrix_unscaled} shows them unscaled, Fig. \ref{SensitivityMatrix_scaled} shows them scaled by the norm of each column of $S$, so that their shapes may be more easily compared. We see that many of the parameters give sensitivity curves of similar shapes - the effects of perturbing any one of these parameters all look similar in terms of model output.

 It is unsurprising then that some of our initial parameter estimates are very loose - this result suggests there is a family of parameters all of which, in terms of the model output we have available, cannot be resolved, and as such that we should view estimates of these parameters with extreme caution. Note that the sensitivity curve that looks least similar to the others in Fig. \ref{SensitivityMatrix_scaled} - $\mu$ - corresponds to a relatively tightly estimated value in Fig. \ref{InitialResults}.
 \todo[inline]{make more quantitivative with colinaerity idnex}

Our inability to estimate some model parameters may stem from structural problems with the model (often termed \textit{structural identifiability}), in which two different parameter sets can give identical model predictions (eq. \ref{nonidentifiable}) . General methods exist to analyse the structural identifiability of an ODE model \cite{Jimenez-Hornero2008,Grewal1976 } \footnote{A simple example of a structurally non - identifiable model is $y = \beta_{1}\beta_{2}x$, where we are given data $(x,y)$, and asked to estimate parameters $\beta_{1}$ and $\beta_{2}$ - we can see that, in principle, only the product can ever be estimated, a problem no amount of data can fix.}.

 \begin{align}
z(\boldsymbol{\theta}_{1},t) = z(\boldsymbol{\theta}_{2},t) \centernot\implies \theta_{1} = \theta_{2}
\label{nonidentifiable}
\end{align}

Problems may also come from limitations in available data - differing experimental conditions may lift the strong correlations between parameter effects on model output (this is often termed \textit{practical identifiability} \cite{Mclean2012}).

In our case, while the model may still turn out to have general structural problems, we suggest that part of the issue is the nature of the forcing used, and the timescales on which the system responds to it. We hypothesize that the system may have two timescales in it - a fast timescale in which eqs. \ref{eq:s} - \ref{eq:c}, representing the hybridization of the sRNA and mRNA into a stable complex, equilibrate in response to external forcing, and a slower timescale, in eqs. \ref{eq:p}, \ref{eq:g}, in which measured fluorescence changes in response to the forcing. If this is the case, then it may be that our experimental data can only probe the system via the fixed point of the models core, and that the parameters contained within those equations can only work to set the particular fixed point values of $p$ and $g$ that the system tends toward. If this were the case, the particular shape of the models response curve would be primarily determined by the parameters appearing in eqs. \ref{eq:p}, \ref{eq:g}, \ref{eq:z} - namely $\mu$, $\beta$, $f_{s}$ and $\Theta$. If all the information contained in the parameters in eqs.\ref{eq:s} - \ref{eq:c} can only be expressed in the form of the models fixed point, it would not be surprising if there were large spaces of parameters, all giving the same model fixed point, which were indistinguishable.

Fig. \ref{model_output} shows model output for all state variables, using the parameter values giving the lowest error on the \textit{13\_9} dataset, and normalised to lie on the same scale. It demonstrates that, at least in some of the parameter sets initially estimated, this difference in time scales exists. The fixed points of eqs. \ref{eq:s} - \ref{eq:c} affects eqs.  \ref{eq:p}, \ref{eq:g} through the $\beta m +f_{s}\beta c$ term - we would expect that, if the fixed point value of this term is what is relevant in determining model response, there would be some consistency in its value across the parameter sets found. Fig. \ref{fixedpoint_f} shows a scatterplot of model fixed point values against error function value, and demonstrates that, though individual parameter values can be spread across very large ranges, they are correlated in such a way as to give similar model fixed points. In Fig. \ref{fixedpoint_mu}, we also see a strong positive correlation between the fixed point values and the values of $\mu$ ($R = 0.9854$ for the tightly clustered data). Taken together with the relatively tightly constrained values of $\mu$ found in Fig. \ref{InitialResults}, these results suggest that in order to minimize error, the model is effectively varying

\begin{figure}[h]	
	\begin{subfigure}[h]{0.49\textwidth}
    \centering
        \includegraphics[scale = 0.26]{14_7_pred_13_9_data}
        \caption{\textit{14\_7} prediction \textit{13\_9} data}
        \label{SensitivityMatrix_unscaled} 
    \end{subfigure}
    \begin{subfigure}[h]{0.49\textwidth}
    \centering
        \includegraphics[scale = 0.26]{13_9_pred_14_7_data}
        \caption{\textit{13\_9} prediction \textit{14\_7} data}
                \label{SensitivityMatrix_scaled} 
    \end{subfigure}
    \caption{Cross validating data by taking parameter estimates from one dataset, and using them to predict another. Here we take the best parameter set fitted on the \textit{13\_9}, and fit it against the \textit{14\_7} dataset, and vice versa } 
\label{CrossValidation}   
\end{figure}

@@

@@

@@

\clearpage

 \begin{itemize}
 \item discuss estimatbility criteria- uniqeuess and sensitivity
 \item put in some sensitivity analysis results
 \item discuss model fixed points
 \item discuss systematic paramater shift.
 \item discuss cross - validation and predicitive ability of parameter sets found.
 \item suggest two times, work out fixed point for all the solutions found, look at structure of soln,
 \item re run from starting pararms.
 \end{itemize}

@@

>>>>

 We can test the predictive ability of our model by cross validating - taking the parameter values found in the fitting of one dataset, and using them to give model predictions for another. We thus take the parameter values giving the best fit for the  \textit{13\_9} dataset and use them to predict the \textit{14\_7} dataset, and vice versa. Results are shown in Fig.\ref{CrossValidation}.

 \begin{figure}[h!]	
	\begin{subfigure}[h]{0.49\textwidth}
    \centering
        \includegraphics[scale = 0.26]{14_7_pred_13_9_data}
        \caption{\textit{14\_7} prediction \textit{13\_9} data}
        \label{SensitivityMatrix_unscaled} 
    \end{subfigure}
    \begin{subfigure}[h]{0.49\textwidth}
    \centering
        \includegraphics[scale = 0.26]{13_9_pred_14_7_data}
        \caption{\textit{13\_9} prediction \textit{14\_7} data}
                \label{SensitivityMatrix_scaled} 
    \end{subfigure}
    \caption{Cross validating data by taking parameter estimates from one dataset, and using them to predict another. Here we take the best parameter set fitted on the \textit{13\_9}, and fit it against the \textit{14\_7} dataset, and vice versa } 
\label{CrossValidation}   
\end{figure}

Though the predictions are reasonable, they are substantially worse than the predictions for the data they were trained on. The first reason for this may be experiment to experiment variability of parameter values \cite{Hu2015}. The second may be that the model is overfitting.

@@

These problems may stem from structural inadequacies in the model (often termed \textit{structural identifiability}), in which two different parameter sets can give identical model predictions \cite{Jimenez-Hornero2008,Grewal1976 }. If this is the case, no amount of experimental data will allow us to estimate parameters, and we must consider reformulating the model. \footnote{A simple example of a structurally non - identifiable model is $y = \beta_{1}\beta_{2}x$, where we are given data $(x,y)$, and asked to estimate parameters $\beta_{1}$ and $\beta_{2}$ - we can see that, in principle, only the product can ever be estimated, a problem no amount of data can fix.}

Problems may also arise for more practical reasons (\textit{practical identifiability} \cite{Mclean2012}). For example, it is possible that in the experimental regime we operate in, parameter effects may be weak, or highly correlated, but in general this is not true. \footnote{\cite{Beck} gives an example in which a parameter only effects model predictions after several hours, though others will affect it at all timescales - in this example, if we only took data for a few minutes, the parameter would be inestimable, but in principle it is not.} In this case, parameter estimates be may improved by taking data in more varied experimental conditions, and attempting to observe as many components of the model output as possible.

 We may begin to investigate these issues in our model by performing a local sensitivity analysis about one of the solutions found in our initial parameter estimation. We numerically estimate the sensitivity matrix, $S$:

@@

It can be shown that if sensitivity co-efficients are linearly dependent over the range of observation values, the associated parameters cannot be simultaneously estimated \cite{Beck}. Related to this fact, a number of measures have been proposed to asses parameter estimability from the sensitivity matrix \cite{Mclean2012}, the simplest of which is to simply plot the sensitivity co-effecients as a function of time, and visually check for obvious linear relations between the curves.

Fig. \ref{SensitivityMatrix} shows plots of each column of $S$, evaluated at the parameter set giving the lowest error in the \textit{13\_9} dataset - these show the evolution of the sensitivity matrix over time.

@@

 Fig. \ref{SensitivityMatrix_unscaled} shows them unscaled, Fig. \ref{SensitivityMatrix_scaled} shows them scaled by the norm of each column of $S$, so that their shapes may be more easily compared. We see that many of the parameters give sensitivity curves of similar shapes - the effects of perturbing any one of these parameters all look similar in terms of model output.

This may help to explain why some of our initial parameter estimates are very loose - this result suggests there is a family of parameters all of which, in terms of the model output we have available, cannot be resolved. As such, we should view estimates of these parameters with extreme caution. Note that the sensitivity curve that looks least similar to the others in Fig. \ref{SensitivityMatrix_scaled} - $\mu$ - corresponds to a relatively tightly estimated value in Fig. \ref{InitialResults}.

We suggest that part of the issue is the step function forcing used in model predictions, and the timescales on which the system responds to it. We hypothesize that the system may have two timescales in it - a fast timescale in which eqs. \ref{eq:s} - \ref{eq:c}, representing the hybridization of the sRNA and mRNA into a stable complex, equilibrate in response to external forcing, and a slower timescale, in eqs. \ref{eq:p}, \ref{eq:g}, in which measured fluorescence changes in response to the forcing. If this is the case, then it may be that our experimental data can only probe the system via the fixed point of eqs. \ref{eq:s} - \ref{eq:c} , and that the parameters contained within those equations can only work to set the particular fixed point values of $p$ and $g$ that the system tends toward. If this were the case, the particular shape of the models response curve would be primarily determined by the parameters appearing in eqs. \ref{eq:p}, \ref{eq:g}, \ref{eq:z} - namely $\mu$, $\beta$, $f_{s}$ and $\Theta$. If all the information contained in the parameters in eqs.\ref{eq:s} - \ref{eq:c} can only be expressed in the form of the models fixed point, it would not be surprising if there were large spaces of parameters, all giving the same model fixed point, which were indistinguishable.

Fig. \ref{model_output} shows model output for all state variables, using the parameter values giving the lowest error on the \textit{13\_9} dataset, and normalised to lie on the same scale.

@@

It demonstrates that, at least in some of the parameter sets initially estimated, this difference in time scales exists. The fixed points of eqs. \ref{eq:s} - \ref{eq:c} affects eqs.  \ref{eq:p}, \ref{eq:g} through the $\beta m +f_{s}\beta c$ term - we would expect that, if the fixed point value of this term is what is relevant in determining model response, there would be some consistency in its value across the parameter sets found. Fig. \ref{fixedpoint_f} shows a scatterplot of model fixed point values against error function value, and demonstrates that, though individual parameter values can be spread across very large ranges, they are correlated in such a way as to give similar model fixed points.

@@

In Fig. \ref{fixedpoint_mu}, we also see a strong positive correlation between the fixed point values and the values of $\mu$ ($R = 0.9854$ for the visually tightly clustered data). Taken together with the relatively tightly constrained values of $\mu$ found in Fig. \ref{InitialResults}, these results suggest that in order to minimize error, the algorithm is effectively trying to vary $\beta m +f_{s}\beta c$ and $\mu$, and finding a `trench' of highly correlated values, which is shown in Fig. \ref{fixedpoint_mu}.

These results further suggest that, while we are trying to minimise model error over our initial high dimensional space, we are effectively working in a lower dimensional space, in which one dimension is the value of the models $\beta m +f_{s}\beta c$ .

 \begin{itemize}
 \item explore degeneracies in fixed point equations
 \item discuss estimatbility criteria- uniqeuess and sensitivity
 \item put in some sensitivity analysis results
 \item discuss model fixed points
 \item discuss systematic paramater shift.
 \item discuss cross - validation and predicitive ability of parameter sets found.
 \item suggest two times, work out fixed point for all the solutions found, look at structure of soln,
 \item re run from starting pararms.
 \end{itemize}

@@

 \begin{itemize}
 \item discuss evidence for model being very over parametrized, and suggest a simpler model may work equally well - suggest biologically interesting parameters may be unobservable.
 \item discuss general methodological flaws - throwing marbles in the air, local minima, is MLE even any good? say data shows even the EA doesn't avoid local minima. The space is so unconstrained. the algorithm can't be expected to just fix all these problems.
 \item suggest additional forcing time data may not help due to two timing nature, 
 \item suggest there MAY be other ways to fit the data, but the wide range of parameter bounds mean we just don't know - the landscape is complicated and huge, and there may be minima widely separated in the parameter space. whose to say which is right?
 \item Further work - on experimnetal side, propoer lit review to rightly constrain some values, fix more. More state variables! should easily be able to fix mu, theta, maybe others too? other forcing functions?
 \item a proper bloody methodology - gaussian processes.
 \item mention this is how sys bio papers do it, and we were given some earlier fitting work which we think is now crap.
 \item suggest a general structural analysis be carried out.

 \end{itemize}
